Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-02-26 TEBenchmarkPaper/article.tex
Consistent citations

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
on this survey (the latter is Colton's own system). We extend this analysis to
some more recent systems in the remaining rows (based on our understanding of
their function): QuickSpec, Speculate~\cite{braquehais2017speculate}, IsaCoSy
and IsaScheme.

@@

\begin{quote}
...``in his subsequent design for an Analytical Engine Mr. Babbage has
shown that material machinery is capable, in theory at least, of
rivalling the labours of the most practised mathematicians in all
branches of their science.'' \cite[p. 498]{jevons}
\end{quote}

The first automated mathematical system, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues \cite{d'ocagne}. Subsequent early systems include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees \cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable \cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
systems were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

@@

% 1. history of terminology
Automated theory formation derives its terminology from psychology in which the
term ``concept formation'' is used (see, for example, \cite{bruner:67}) to
describe the search for features which differentiate exemplars from
non-exemplars of various categories. Lenat used this term in his 1977 paper:
{\em Automated Theory Formation in Mathematics} \cite{lenat:77}.

% 4. example systems - AM, HR, ...
When Lenat built the AM system \cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in \cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral
\cite{buchanan:75}. No system could perform both of these tasks: Lenat
saw this as the next step. AM was designed to both construct new
concepts and conjecture relationships between them; fully automating
the cycle of discovery in mathematics. Lenat describes this as
follows:

\begin{quote}
``What we are describing is a computer program which
defines new concepts, investigates them, notices
regularities in the data about them, and conjectures
relationships between them. This new information is used
by the program to evaluate the newly-defined concepts,
concentrate upon the most interesting ones, and iterate the
entire process.'' \cite[p. 834]{lenat:77}
\end{quote}

AM was a rule-based system which used a frame-like scheme to represent
its knowledge, enlarged its knowledge base via a collection of
heuristic rules, and controlled the firing of these rules via an
agenda mechanism. Lenat chose elementary arithmetic as the development
domain because he could use personal introspection for the heuristics
for constructing and evaluating concepts. Given the age of this
discipline, Lenat thought it unlikely that AM would make significant
discoveries, although he did cite its ``ultimate achievements'' as the
concepts and conjectures it discovered (or could have discovered). He
suggested various criteria by which his system could be evaluated,
many of which focused on an exploration of the techniques. For
instance, he considered generality (running AM in new domains) and how
finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). %most of which were qualitative.
Lenat saw his system and future developments in this field as having
implications for mathematics itself (finding results of significance),
for automating mathematics research (developing AI techniques), and
for designing ``scientist assistant'' programs (aids for
mathematicians). This shows a broad spread of motivation. Despite the
seeming success of the AM system, it is one of the most criticised
pieces of AI research. In their case study in methodology, Ritchie and
Hanna analysed Lenat's written work on AM and found that there was a
large discrepancy between his theoretical claims and the implemented
program \cite{partridge}. For instance, Lenat made claims about how AM
invented natural numbers from sets, whereas it used one heuristic
which was specifically written in order to make this connection (and
not used in any other context). Another problem was that the processes
were sometimes under-explained. For an argument of why many of the
claims made by Lenat about AM were false, see chapter 13 of
\cite{colton:book}.

@@

The phrase Mathematical Theory Exploration (MTE) is a recent term
which seems to have originated with Buchberger and colleagues (see,
for example, \cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm \cite{buchberger:04}.

Recent systems developed in this area include MATHsAiD \cite{roy},
IsaScheme \cite{MontanoRivas2011} and IsaCosy \cite{johansson}. The
goal of the MATHsAiD (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions) project is to build a tool which
takes in a set of axioms, concept definitions and a logic and applies
its inference rules to reason from the axioms to theorems. The
motivation is to produce a tool which will help mathematicians to
explore the consequences of a set of axioms or a particular concept
definition. One of the main challenges of the project has been to
automatically evaluate interestingness: to distinguish important
theorems from results which, although they follow from a set of
axioms, are of little mathematical interest.

@@

\begin{align*}
0*y  &= 0 \\
suc(x) * y &=y + (x*y).
\end{align*}
The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from
\cite{MontanoRivas2011}).

%IsaCosy
The IsaCosy system (Isabelle Conjecture Synthesis) is a program for
inductive theory formation, which synthesises conjectures from the
available constants and free variables. Only terms which do not
contain subterms that can be rewritten by a set of rewrite rules can
be synthesised.  Conjectures are tested by sending them to a
counterexample checker and, if no counterexamples are found, then sent
to IsaPlanner which attempts to prove them.

@@

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to \cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and current
systems such as MATHsAiD \cite{roy}, IsaScheme \cite{MontanoRivas2011}
and IsaCosy \cite{johansson} are bridging the methodological gap
between ATF and MTE. This is particularly true in terms of
fully-automatic/user-interactive and the domain-specific/general
aspects.

@@

If we limit ourselves to exploring higher-order theories with inductive types
(a domain closely matching our interest in analysing software), we do find a
direct comparison of three systems~\cite{claessen2013automating}: HipSpec,
IsaCoSy and IsaScheme. QuickSpec, shown in table \ref{table:colton}, is the
conjecture-generating component of HipSpec (which sends those conjectures into
automated theorem provers).

@@

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as
possible\cite{pelletier2002development}. These problem sets are an opportunity
for theory exploration, as their definitions and theorems can be used as a
corpus, in the same way that Isabelle libraries have been used in the past.

>>>>

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
    to include those from~\cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec) and the similar Speculate
    system~\cite{braquehais2017speculate}. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
on this survey (the latter is Colton's own system). We extend this analysis to
some more recent systems in the remaining rows (based on our understanding of
their function): QuickSpec~\cite{QuickSpec},
Speculate~\cite{braquehais2017speculate},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and IsaScheme~\cite{MontanoRivas2011}.

@@

\begin{quote}
...``in his subsequent design for an Analytical Engine Mr. Babbage has
shown that material machinery is capable, in theory at least, of
rivalling the labours of the most practised mathematicians in all
branches of their science.''~\cite[p. 498]{jevons}
\end{quote}

The first automated mathematical system, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early systems include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
systems were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

@@

% 1. history of terminology
Automated theory formation derives its terminology from psychology in which the
term ``concept formation'' is used (see, for example,~\cite{bruner:67}) to
describe the search for features which differentiate exemplars from
non-exemplars of various categories. Lenat used this term in his 1977 paper:
{\em Automated Theory Formation in Mathematics}~\cite{lenat:77}.

% 4. example systems - AM, HR, ...
When Lenat built the AM system~\cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in~\cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
system could perform both of these tasks: Lenat saw this as the next step. AM
was designed to both construct new concepts and conjecture relationships between
them; fully automating the cycle of discovery in mathematics. Lenat describes
this as follows:

\begin{quote}
``What we are describing is a computer program which
defines new concepts, investigates them, notices
regularities in the data about them, and conjectures
relationships between them. This new information is used
by the program to evaluate the newly-defined concepts,
concentrate upon the most interesting ones, and iterate the
entire process.''~\cite[p. 834]{lenat:77}
\end{quote}

AM was a rule-based system which used a frame-like scheme to represent
its knowledge, enlarged its knowledge base via a collection of
heuristic rules, and controlled the firing of these rules via an
agenda mechanism. Lenat chose elementary arithmetic as the development
domain because he could use personal introspection for the heuristics
for constructing and evaluating concepts. Given the age of this
discipline, Lenat thought it unlikely that AM would make significant
discoveries, although he did cite its ``ultimate achievements'' as the
concepts and conjectures it discovered (or could have discovered). He
suggested various criteria by which his system could be evaluated,
many of which focused on an exploration of the techniques. For
instance, he considered generality (running AM in new domains) and how
finely-tuned various aspects of the program are (the agenda, the
interaction of the heuristics, etc). %most of which were qualitative.
Lenat saw his system and future developments in this field as having
implications for mathematics itself (finding results of significance),
for automating mathematics research (developing AI techniques), and
for designing ``scientist assistant'' programs (aids for
mathematicians). This shows a broad spread of motivation. Despite the
seeming success of the AM system, it is one of the most criticised
pieces of AI research. In their case study in methodology, Ritchie and
Hanna analysed Lenat's written work on AM and found that there was a
large discrepancy between his theoretical claims and the implemented
program~\cite{partridge}. For instance, Lenat made claims about how AM
invented natural numbers from sets, whereas it used one heuristic
which was specifically written in order to make this connection (and
not used in any other context). Another problem was that the processes
were sometimes under-explained. For an argument of why many of the
claims made by Lenat about AM were false, see chapter 13 of~\cite{colton:book}.

@@

The phrase Mathematical Theory Exploration (MTE) is a recent term
which seems to have originated with Buchberger and colleagues (see,
for example,~\cite{buchberger}). His motivation is to support
mathematicians during their exploration of mathematical theories. This
support is intended to be for the straightforward reasoning, which he
argues, covers most mathematical thought, rather than the ingenious
points, which he leaves for human mathematicians. Buchberger's long
term goal is to provide routine tools for the exploration activity of
working mathematicians, to support the invention and structured
build-up of mathematical knowledge. The Theorema project aims at
prototyping features of a system for such a purpose. These features
include Integration of the Functionality of Current Mathematical
Systems (retention of the full power of current numerics and computer
algebra systems, as well as enabling the user to add their own
algorithms to the system); Attractive Syntax (input and output is
readable and presented attractively, and can be personalised by the
user); and Structured Mathematical Knowledge Bases (tools are provided
for building and using large mathematical knowledge
libraries). Buchberger has evaluated the potential of this strategy by
illustrating the automated synthesis of his own Gr\"obner bases
algorithm~\cite{buchberger:04}.

Recent systems developed in this area include MATHsAiD~\cite{roy},
IsaScheme~\cite{MontanoRivas2011} and
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}. The goal of the
MATHsAiD (Mechanically Ascertaining Theorems from Hypotheses, Axioms and
Definitions) project is to build a tool which takes in a set of axioms, concept
definitions and a logic and applies its inference rules to reason from the
axioms to theorems. The motivation is to produce a tool which will help
mathematicians to explore the consequences of a set of axioms or a particular
concept definition. One of the main challenges of the project has been to
automatically evaluate interestingness: to distinguish important theorems from
results which, although they follow from a set of axioms, are of little
mathematical interest.

@@

\begin{align*}
0*y  &= 0 \\
suc(x) * y &=y + (x*y).
\end{align*}
The new multiplication function $f$ can itself be used to instantiate
variables in the same scheme, resulting in the invention of the
exponentiation concept (these examples are taken from~\cite{MontanoRivas2011}).

%IsaCoSy
The IsaCoSy system (Isabelle Conjecture Synthesis) is a program for
inductive theory formation, which synthesises conjectures from the
available constants and free variables. Only terms which do not
contain subterms that can be rewritten by a set of rewrite rules can
be synthesised.  Conjectures are tested by sending them to a
counterexample checker and, if no counterexamples are found, then sent
to IsaPlanner which attempts to prove them.

@@

The different backgrounds of the people who named the fields
``automated theory formation'' (Lenat) and ``mathematical theory
exploration'' (Buchberger) perhaps reflects different metaphysical
perspectives on invention (in which case new mathematical ideas are
being {\em formed}, created or produced) and discovery (in which case
abstract mathematical objects exist independently of us and are {\em
  explored} or investigated) in mathematics.\footnote{There is a vast
  literature on the Platonic and the constructivist view of
  mathematics: we shall not venture down this path here (interested
  readers are referred to~\cite{hersh:97,shapiro}). Instead, for now,
  we make the pragmatic assumption that while there may well be
  different cognitive processes involved in invention than those
  involved in discovery, the fields of ATF and MTE are not currently
  concerned with this level of detail and so the philosophical
  distinction is not relevant for our purposes.} As can be seen above,
some of these historical differences are disappearing, and current
systems such as MATHsAiD~\cite{roy}, IsaScheme~\cite{MontanoRivas2011}
and IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation} are bridging the
methodological gap between ATF and MTE. This is particularly true in terms of
fully-automatic/user-interactive and the domain-specific/general
aspects.

@@

If we limit ourselves to exploring higher-order theories with inductive types
(a domain useful for analysing software), we do find a direct comparison of
three systems~\cite{claessen2013automating}: HipSpec, IsaCoSy and IsaScheme.
QuickSpec, shown in table \ref{table:colton}, is the conjecture-generating
component of HipSpec.

@@

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as
possible~\cite{pelletier2002development}. These problem sets are an opportunity
for theory exploration, as their definitions and theorems can be used as a
corpus, in the same way that Isabelle libraries have been used in the past.
